Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 6: Used semicolons instead of commas. Added line breaks. Fixed some typos.

Six papers were categorized under the theme of Exploring Sustainability. The
group included four of the authors from the six papers submitted and an
additional number of participants who expressed an interest in the theme at the
workshop. Each paper had a different perspective on the concept of
sustainability, which ranged from the sustainment of communities to defining
sustainability as a first-class, composite non-functional requirement.
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Discussion and Actions}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsubsection{Discussion}

Each author was invited to outline the key action from their paper as a
potential discussion point for the group; where the authors were not present,
the group facilitators outlined the actionable points from their papers. The key
actions from the six papers included recommendations for improving practice in
sustainable software engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs},
the development of Software and Infrastructure as a Service as a mechanism for
fostering sustainable science communities~\cite{wssspe2_patra}, developer
incentives for code contributions to open source projects~\cite{wssspe2_pierce},
the establishment of a set software engineering principles based in scalability,
reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software
architectures as a mechanism for architectural-level reasoning about
sustainability~\cite{wssspe2_venters}. The group took the position of viewing
sustainability from the perspective of addressing the challenges related to the
development, deployment, and maintenance of reusable
software~\cite{katzJORS2014}. The principal focus of the initial discussion
considered how to foster cultural change towards developing sustainable software
in academic environments. It was suggested that a new requirement driven by the
agencies funding research projects where software was an intrinsic part of
enabling the research program would be an additional element within the grant
proposal to provide a sustainability plan for sustaining the software. This type
of initiative would provide the necessary incentive and motivation for
researchers to consider how to sustain their software beyond the lifetime of the
project.

The discussion then focused on a need for a common language and a definition on
the concept of sustainability that moved beyond the current fuzzy definitions
where time was the simple measurement of sustainment. It was suggested that
there was a need to identify tangible actions that underpin sustainability that
developers could incorporate into the development stream of their software. This
prompted a debate regarding whether sustainability should be considered as a
non-functional requirement or software quality as defined within ISO/IEC
25010~\cite{iso25010}. The focus of the discussion was based on the paper by
Venters et. al.~\cite{wssspe2_venters}, that suggested that sustainability is a
first-class, composite, non-functional requirement composed of a number of
sub-characteristics. It was generally agreed that maintainability and
extensibility were key qualities underlying sustainability. In addition, the
group also discussed what other non-functional requirements would contribute to
the development of sustainable software e.g. reusability and scalability.
However, it was recognized that there was a need to identify appropriate metrics
and measures.

The group also discussed whether the concept of sustainability itself was a
barrier to achieving sustainable software. It was suggested that ``the first
rule of software sustainability is do not to talk about software
sustainability.'' Instead there should be a focus on best software engineering
practice. Playing devil's advocate, it was asked that if the focus on the
concept of sustainability was ignored, what current software engineering
practices and principles could be utilized by software developers and domain
scientists to achieve sustainability? This raised the question of why existing
software engineering knowledge, such as that contained in
SWEBOK~\cite{Bourque2014}, was largely ignored and to what extent the
environment has a strong influence on practice within the scientific and
engineering community. As a result, there is a need to identify best practices
and reach a consensus beyond the theories that underpin the discipline of
software engineering. Similarly, how could we translate or distill some of the
key building blocks that underpin software engineering practice?

>>>>

Six papers were categorized under the theme of Exploring Sustainability. The
group included four of the authors from the six papers submitted and an
additional number of participants who expressed an interest in the theme at the
workshop. Each paper had a different perspective on the concept of
sustainability, which ranged from the sustainment of communities to defining
sustainability as a first-class, composite non-functional requirement.

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Discussion and Actions}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsubsection{Discussion}

Each author was invited to outline the key action from their paper as a
potential discussion point for the group; where the authors were not present,
the group facilitators outlined the actionable points from their papers. The key
actions from the six papers included recommendations for improving practice in
sustainable software engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs};
development of \emph{Software and Infrastructure as a Service} as a mechanism for
fostering sustainable science communities~\cite{wssspe2_patra}; developer
incentives for code contributions to open source projects~\cite{wssspe2_pierce};
establishment of a set of software engineering principles based on scalability,
reproducibility, and energy efficiency~\cite{wssspe2_shi}; and applying software
architectures as a mechanism for architectural-level reasoning about
sustainability~\cite{wssspe2_venters}. The group took the position of viewing
sustainability from the perspective of addressing the challenges related to the
development, deployment, and maintenance of reusable
software~\cite{katzJORS2014}.

The principal focus of the initial discussion
considered how to foster cultural change towards developing sustainable software
in academic environments. It was suggested that a new requirement driven by the
agencies funding research projects where software was an intrinsic part of
enabling the research program would be an additional element within the grant
proposal to provide a sustainability plan for sustaining the software. This type
of initiative would provide the necessary incentive and motivation for
researchers to consider how to sustain their software beyond the lifetime of the
project.

The discussion then focused on a need for a common language and a definition on
the concept of sustainability that moved beyond the current fuzzy definitions
where time was the simple measurement of sustainment. It was suggested that
there was a need to identify tangible actions that underpin sustainability that
developers could incorporate into the development stream of their software. This
prompted a debate regarding whether sustainability should be considered as a
non-functional requirement or software quality as defined within ISO/IEC
25010~\cite{iso25010}. The focus of the discussion was based on the paper by
Venters et al.~\cite{wssspe2_venters}, which suggested that sustainability is a
first-class, composite, non-functional requirement composed of a number of
sub-characteristics. It was generally agreed that maintainability and
extensibility were key qualities underlying sustainability. In addition, the
group also discussed what other non-functional requirements would contribute to
the development of sustainable software, e.g., reusability and scalability.
However, it was recognized that there was a need to identify appropriate metrics
and measures.

The group also discussed whether the concept of sustainability itself was a
barrier to achieving sustainable software. It was suggested that ``the first
rule of software sustainability is do not talk about software
sustainability.'' Instead there should be a focus on best software engineering
practice. Playing devil's advocate, it was asked that if the focus on the
concept of sustainability was ignored, what current software engineering
practices and principles could be utilized by software developers and domain
scientists to achieve sustainability? This raised the question of why existing
software engineering knowledge, such as that contained in
SWEBOK~\cite{Bourque2014}, was largely ignored and to what extent the
environment has a strong influence on practice within the scientific and
engineering community. As a result, there is a need to identify best practices
and reach a consensus beyond the theories that underpin the discipline of
software engineering. Similarly, how could we translate or distill some of the
key building blocks that underpin software engineering practice?
